The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic

### 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Started working on replication of SWN

@@

Given the heavy human and economic cost of the pandemic, there has been
a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research has even called
this deluge of publications a ``carnage of substandard research''
{[}5{]}. Part of the challenge of maintaining quality standards in
published research is that, despite an abundance of recommendations and
guidelines {[}6--9{]}, in practice reproducibility has remained a lofty
but somewhat aspirational goal {[}10,11{]}. As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic {[}12,13{]}, a situation that
does not appear to have changed substantially since {[}14,15{]}.

The push for open data and software, along with more strenuous efforts
towards open, reproducible research, is simply a continuation of
long-standing scientific practices of independent verification. Despite
the (at times disproportionate) attention that high profile scandals in
science tend to elicit in the media, science as a collective endeavor is
remarkable for being a self-correcting enterprise, one with built-in
mechanisms and incentives to weed out erroneous ideas. Over the long
term, facts tend to prevail in science. At stake is the shorter-term
impacts that research may have in other spheres of economic and social
life. The case of economists Reinhart and Rogoff comes to mind: by the
time the inaccuracies in their research were uncovered {[}see 16{]},
their claims about debt and economic growth had already been seized by
policy-makers on both sides of the Atlantic to justify austerity
policies in the aftermath of the Great Recession of
2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
  ``Reinhart--Rogoff may have had more immediate influence on public
  debate than any previous paper in the history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years {[}17{]}.

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic {[}18{]}. The fact that the worst of the pandemic was initially
felt in dense population centers such as Wuhan, Milan, Madrid, and New
York, prompted a flurry of research into the associations between
density and the spread of the pandemic. Some important questions hang on
the results of these research efforts. For example, are lower density
regions safer from the pandemic? Are de-densification policies
warranted, at least in the short term? And in the longer term, will the
risks of life in high density regions presage a flight from cities? Over
the past year, numerous papers have sought to throw light into the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

@@

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density may have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density {[}see for example the research on urban fields and
time-geography 20,21,22{]}. Mathematically, models of exposure and
contagion indicate that higher densities can catalyze the transmission
of contagious diseases {[}23,24{]}. Models such as these were likely at
the root of messages, by some figures in positions of authority, that
low density regions faced lower risks from the pandemic\footnote{Governor
  Kristi Noem of South Dakota, for example, claimed that sparse
  population density allowed her state to face the pandemic without the
  need for strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
contagion are valid at small-to-medium spatial scales (and presumably,
small temporal scales too, such as time spent in restaurants, concert
halls, cruises), and the results do not necessarily transfer to larger
spatial units and different time scales. There are good reasons for
this: while in a restaurant, one can hardly avoid being in proximity to
other customers-however, a person can choose to (or be forced to as a
matter of policy) not go to a restaurant in the first place.
Nonetheless, the idea that high density correlates with high
transmission is so intuitive that it is often taken for granted even at
larger scales {[}e.g., 25,26{]}. At larger scales, however, there exists
the possibility of behavioral adaptations, which are difficult to
capture in the mechanistic framework of differential equations {[}or can
be missing in agent-based models, 27{]}; these adaptations, in fact, can
be a key aspect of disease transmission.

A plausible behavioral adaptation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk {[}28--30{]}. In the case of
COVID-19, Chauhan et al. {[}31{]} have found that perception of risks in
the US varies between rural, suburban, and urban residents, with rural
residents in general displaying less concerns about the virus. It is
possible that people who listened to the message of leaders saying that
they were safe because of low density may not have taken adequate
precautions against the virus. People in dense places who could more
directly observe the impact of the pandemic may have become overly
cautious. Both Paez et al. {[}32{]} and Hamidi et al. {[}33{]} posit
this mechanism (i.e., greater compliance with social distancing in
denser regions) to explain the results of their analyses. The evidence
available does indeed show that there were important changes in behavior
during the pandemic, at least with respect to mobility {[}34--36{]};
furthermore, shelter in place orders may have had greater buy-in from
the public in higher density regions {[}37{]}, and the behavior may have
persisted beyond the duration of official social-distancing policies
{[}38{]}. In addition, there is evidence that changes in mobility
correlated with the trajectory of the pandemic {[}39,40{]}. Given the
potential for behavioral adaptation, the question of density becomes
more nuanced: it is not just a matter of proximity, but also of human
behavior, which is better studied using population-level data and
models.

@@

On the one hand, there are studies that report positive associations
between population density and various COVID-19-related outcomes. Bhadra
{[}41{]}, for example, reported a moderate positive correlation between
the spread of COVID-19 and population density at the district level in
India, however their analysis was bivariate and did not control for
other variables, such as income. Similarly, Kadi and Khelfaoui {[}42{]}
found a positive and significant correlation between number of cases and
population density in cities in Algeria in a series of simple regression
models (i.e., without other controls). A question in these relatively
simple analyses is whether density is not a proxy for other factors.
Other studies have included controls, such as Pequeno et al. {[}43{]}, a
team that reported a positive association between density and cumulative
counts of confirmed COVID-19 cases in state capitals in Brazil after
controlling for covariates, including income, transport connectivity,
and economic status. In a similar vein, Fielding-Miller et al. {[}44{]}
reported a positive relationship between the absolute number of COVID-19
deaths and population density (rate) in rural counties in the US. Roy
and Ghosh {[}45{]} used a battery of machine learning techniques to find
discriminatory factors, and a positive and significant association
between COVID-19 infection and death rates in US states. Wong and Li
{[}46{]} also found a positive and significant association between
population density and number of confirmed COVID-19 cases in US
counties, using both univariate and multivariate regressions with
spatial effects. Most recently, Sy et al. {[}19{]} reported that the
basic reproductive number of COVID-19 in US counties tends to increase
with population density, but at a decreasing rate at higher densities.

On the gripping hand, a number of studies report non-significant or
negative associations between population density and COVID-19 outcomes.
This includes the research of Sun et al. {[}47{]} who did not find
evidence of significant correlation between population density and
confirmed number of cases per day \emph{in conditions of lockdown} in
China. This finding echoes the results of Paez et al. {[}32{]}, who in
their study of provinces in Spain reported non-significant associations
between population density and infection rates in the early days of the
first wave of COVID-19, and negative significant associations in the
later part of the first lockdown. Similarly, {[}48{]} found zero or
negative associations between population density and infection
numbers/deaths by country. Fielding-Miller et al. {[}44{]} found a
negative relationship between COVID-19 deaths and population density
urban counties in the US. In their investigation of doubling time, White
and Hébert-Dufresne {[}49{]} identified a negative and significant
correlation between population density and doubling time in US states.
Likewise, {[}50{]} fond a small negative (and significant) association
between population density and COVID-19 morbidity in districts in
Tehran. And two of the most complete studies in the US by Hamidi et al.
{[}33,51{]} used an extensive set of controls to find negative and
significant correlations between density at the level of counties in the
US and COVID-19 cases and fatalities.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in the following sections), the study of Sy et
al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
of a researcher with expertise in spatial modelling. Such an expert
would likely ask some of the following questions: how were missing
counties treated? Is it possible to spatially interpolate missing
observations? What are the implications of the spatial sampling
framework used in the analysis? Was there evidence of spatial
autocorrelation in the residuals of the models? These are questions that
in most cases would not occur to a researcher who has not been exposed
to spatial statistics or spatial econometrics. Nonetheless, they are
relevant and important. Fortunately, SWN give an example of a reasonably
open, reproducible research product: their paper is accompanied by (most
of) the data and (most of) the code used in the analysis. This means
that an independent expert can, with only a moderate investment of time
and effort, replicate the results in the paper, as well as ask
additional questions.

Alas, reproducibility is not necessarily the norm.

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons {[}e.g., 52{]}. But
in many other cases the data are publicly available, as has been
commonly the case with population-level COVID-19 information. Often the
provenance of the data is documented, but the data themselves are not
shared {[}25,33,41,44,51,e.g., 53,54--56{]}. As any researcher can
attest, whether a graduate student or a seasoned scientist, collecting,
organizing, and preparing data for a project can take a substantial
amount of time. Pointing to the sources of data, even when these sources
are public, is a small step towards reproducibility-but only a very
small one. Faced with the prospect of having to recreate a data set from
raw sources is probably sufficient to dissuade all but the most
dedicated (or stubborn) researcher. This is true even if part of the
data are shared {[}e.g., 46{]}. In other cases, data are shared, but the
processes to document the preparation of the data are not fully
documented {[}48,e.g., 57{]}. These processes matter, as shown by the
errors in the spreadsheets of Reinhart and Rogoff {[}16{]}, and the data
of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis {[}43,e.g., 58,59{]}. Making code
available only ``on demand'' {[}e.g., 60{]} is an unnecessary barrier
when most journals offer the facility to share supplemental materials
online. Then there are those papers that strive towards reproducibility,
sharing well-documented processes and data, as well as the code used in
any analyses reported {[}e.g., 32,37,49,61{]}.

@@

Fit (mixed) linear models as in Sy et al.:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log , }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##        AIC      BIC    logLik
##   2383.621 2403.808 -1187.811
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:    0.165504 0.6648325
## 
## Fixed effects:  R ~ density_log 
##                 Value  Std.Error   DF  t-value p-value
## (Intercept) 2.2740045 0.05468629 1099 41.58272       0
## density_log 0.1621127 0.01485543 1099 10.91269       0
##  Correlation: 
##             (Intr)
## density_log 0.797 
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.3847932 -0.5972261 -0.1682299  0.3681864 16.2716371 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                 lower      est.     upper
## (Intercept) 2.1667032 2.2740045 2.3813058
## density_log 0.1329645 0.1621127 0.1912609
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                     lower     est.     upper
## sd((Intercept)) 0.1080424 0.165504 0.2535261
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6375478 0.6648325 0.6932849
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private , }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##        AIC      BIC    logLik
##   2385.478 2410.707 -1187.739
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:   0.1358322 0.6649714
## 
## Fixed effects:  R ~ density_log + private 
##                 Value Std.Error   DF   t-value p-value
## (Intercept)  3.347070 0.3418910 1098  9.789874  0.0000
## density_log  0.145258 0.0156188 1098  9.300177  0.0000
## private     -0.012547 0.0039434 1098 -3.181760  0.0015
##  Correlation: 
##             (Intr) dnsty_
## density_log -0.222       
## private     -0.988  0.344
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.7347592 -0.6005758 -0.1610880  0.3876853 16.2104980 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                   lower       est.        upper
## (Intercept)  2.67623619  3.3470697  4.017903160
## density_log  0.11461177  0.1452579  0.175904019
## private     -0.02028416 -0.0125468 -0.004809444
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                      lower      est.    upper
## sd((Intercept)) 0.08059718 0.1358322 0.228921
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6376445 0.6649714 0.6934695
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model3 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome, }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model3)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##       AIC     BIC   logLik
##   2393.98 2424.25 -1190.99
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:   0.1373995 0.6651151
## 
## Fixed effects:  R ~ density_log + private + hincome 
##                 Value Std.Error   DF   t-value p-value
## (Intercept)  3.385550 0.3933810 1097  8.606289  0.0000
## density_log  0.146826 0.0171534 1097  8.559585  0.0000
## private     -0.012707 0.0040548 1097 -3.133786  0.0018
## hincome     -0.003256 0.0151105 1097 -0.215493  0.8294
##  Correlation: 
##             (Intr) dnsty_ privat
## density_log  0.027              
## private     -0.950  0.211       
## hincome     -0.493 -0.412  0.228
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.7162603 -0.5984817 -0.1626793  0.3887931 16.2029477 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model3)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                   lower         est.       upper
## (Intercept)  2.61368596  3.385550103  4.15741424
## density_log  0.11316880  0.146825984  0.18048317
## private     -0.02066315 -0.012707018 -0.00475089
## hincome     -0.03290488 -0.003256195  0.02639249
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                    lower      est.     upper
## sd((Intercept)) 0.081461 0.1373995 0.2317504
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6377582 0.6651151 0.6936455
\end{verbatim}

Linear mixed models of R0 and density + \% private transportation +
median income + interaction of density and \% private transportation

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model4 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log}\SpecialCharTok{*}\NormalTok{private }\SpecialCharTok{+}\NormalTok{ hincome, }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model4)}
\end{Highlighting}
\end{Shaded}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Given the heavy human and economic cost of the pandemic, there has been
a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research has even called
this deluge of publications a ``carnage of substandard research''
{[}5{]}. Part of the challenge of maintaining quality standards in
published research is that, despite an abundance of recommendations and
guidelines {[}6--9{]}, in practice reproducibility has remained a lofty
and somewhat aspirational goal {[}10,11{]}. As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic {[}12,13{]}, and the situation
does not appear to have changed substantially since {[}14,15{]}.

The push for open data and software, along with more strenuous efforts
towards open, reproducible research, is simply a continuation of
long-standing scientific practices of independent verification. Despite
the (at times disproportionate) attention that high profile scandals in
science tend to elicit in the media, science as a collective endeavor is
remarkable for being a self-correcting enterprise, one with built-in
mechanisms and incentives to weed out erroneous ideas. Over the long
term, facts tend to prevail in science. At stake is the shorter-term
impacts that research may have in other spheres of economic and social
life. The case of economists Reinhart and Rogoff comes to mind: by the
time the inaccuracies and errors in their research were uncovered {[}see
16{]}, their claims about debt and economic growth had already been
seized by policy-makers on both sides of the Atlantic to justify
austerity policies in the aftermath of the Great Recession of
2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
  ``Reinhart--Rogoff may have had more immediate influence on public
  debate than any previous paper in the history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years {[}17{]}.

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic {[}18{]}. The fact that the worst of the pandemic was initially
felt in dense population centers such as Wuhan, Milan, Madrid, and New
York, brought a torrent of research into the associations between
density and the spread of the pandemic. Some important questions hang on
the results of these research efforts. For example, are lower density
regions safer from the pandemic? Are de-densification policies
warranted, at least in the short term? And in the longer term, will the
risks of life in high density regions presage a flight from cities? Over
the past year, numerous papers have sought to throw light into the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

@@

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density may have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density {[}see for example the research on urban fields and
time-geography 20,21,22{]}. Mathematically, models of exposure and
contagion indicate that higher densities can catalyze the transmission
of contagious diseases {[}23,24{]}. Models such as these were likely at
the root of messages, by some figures in positions of authority, that
regions with sparse population densities faced lower risks from the
pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
  claimed that sparse population density allowed her state to face the
  pandemic down without the need for strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
contagion are valid at small-to-medium spatial scales (and presumably,
small temporal scales too, such as time spent in restaurants, concert
halls, cruises), and the results do not necessarily transfer to larger
spatial units and different time scales. There are solid reasons for
this: while in a restaurant, one can hardly avoid being in proximity to
other customers-however, a person can choose to (or be forced to as a
matter of policy) not go to a restaurant in the first place.
Nonetheless, the idea that high density correlates with high
transmission is so intuitive that it is often taken for granted even at
larger scales {[}e.g., 25,26{]}. At larger scales, however, there exists
the possibility of behavioral adaptations, which are difficult to
capture in the mechanistic framework of differential equations {[}or can
be missing in agent-based models, e.g., 27{]}; these adaptations, in
fact, can be a key aspect of disease transmission.

A plausible behavioral adaptation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk {[}28--30{]}. In the case of
COVID-19, Chauhan et al. {[}31{]} have found that perception of risks in
the US varies between rural, suburban, and urban residents, with rural
residents in general expressing less concern about the virus. It is
possible that people who listened to the message of leaders saying that
they were safe because of low density may not have taken adequate
precautions against the virus. People in dense places who could more
directly observe the impact of the pandemic may have become overly
cautious. Both Paez et al. {[}32{]} and Hamidi et al. {[}33{]} posit
this mechanism (i.e., greater compliance with social distancing in
denser regions) to explain the results of their analyses. The evidence
available does indeed show that there were important changes in behavior
with respect to mobility during the pandemic {[}34--36{]}; furthermore,
shelter in place orders may have had greater buy-in from the public in
higher density regions {[}37{]}, and the associated behavior may have
persisted beyond the duration of official social-distancing policies
{[}38{]}. In addition, there is evidence that changes in mobility
correlated with the trajectory of the pandemic {[}39,40{]}. Given the
potential for behavioral adaptation, the question of density becomes
more nuanced: it is not just a matter of proximity, but also of human
behavior, which is better studied using population-level data and
models.

@@

On the one hand, there are studies that report positive associations
between population density and various COVID-19-related outcomes. Bhadra
{[}41{]}, for example, reported a moderate positive correlation between
the spread of COVID-19 and population density at the district level in
India, however their analysis was bivariate and did not control for
other variables, such as income. Similarly, Kadi and Khelfaoui {[}42{]}
found a positive and significant correlation between number of cases and
population density in cities in Algeria in a series of simple regression
models (i.e., without other controls). A question in these relatively
simple analyses is whether density is not a proxy for other factors.
Other studies have included controls, such as Pequeno et al. {[}43{]}, a
team that reported a positive association between density and cumulative
counts of confirmed COVID-19 cases in state capitals in Brazil after
controlling for covariates, including income, transport connectivity,
and economic status. In a similar vein, Fielding-Miller et al. {[}44{]}
reported a positive relationship between the absolute number of COVID-19
deaths and population density (rate) in rural counties in the US. Roy
and Ghosh {[}45{]} used a battery of machine learning techniques to find
discriminatory factors, and a positive and significant association
between COVID-19 infection and death rates in US states. Wong and Li
{[}46{]} also found a positive and significant association between
population density and number of confirmed COVID-19 cases in US
counties, using both univariate and multivariate regressions with
spatial effects. More recently, Sy et al. {[}19{]} reported that the
basic reproductive number of COVID-19 in US counties tended to increase
with population density, but at a decreasing rate at higher densities.

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. {[}47{]} who did not find evidence
of significant correlation between population density and confirmed
number of cases per day \emph{in conditions of lockdown} in China. This
finding echoes the results of Paez et al. {[}32{]}, who in their study
of provinces in Spain reported non-significant associations between
population density and infection rates in the early days of the first
wave of COVID-19, and negative significant associations in the later
part of the first lockdown. Similarly, {[}48{]} found zero or negative
associations between population density and infection numbers/deaths by
country. Fielding-Miller et al. {[}44{]} contrast their finding about
rural counties with a negative relationship between COVID-19 deaths and
population density in urban counties in the US. For their part, in their
investigation of doubling time, White and Hébert-Dufresne {[}49{]}
identified a negative and significant correlation between population
density and doubling time in US states. Likewise, {[}50{]} fond a small
negative (and significant) association between population density and
COVID-19 morbidity in districts in Tehran. Finally, two of the most
complete studies in the US {[}by Hamidi et al.; {[}51{]}; {[}33{]}{]}
used an extensive set of controls to find negative and significant
correlations between density and COVID-19 cases and fatalities at the
level of counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in the following sections), the study of Sy et
al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
of a researcher with expertise in spatial modelling.

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. The basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made certain seemingly reasonable modelling
decision, for instance basing their analysis only on counties with valid
observations. An experiences spatial modeler would likely ask some of
the following questions on reading SWN's paper: how were missing
counties treated? Is it possible to spatially interpolate the missing
observations? What are the implications of the spatial sampling
framework used in the analysis? And, was there evidence of spatial
autocorrelation in the residuals of the models? These are questions that
in most cases would not occur to a researcher who has not been exposed
to spatial statistics or spatial econometrics. Nonetheless, they are
relevant and important. Fortunately, SWN give an example of a reasonably
open, reproducible research product: their paper is accompanied by (most
of) the data and (most of) the code used in the analysis. This means
that an independent expert can, with only a moderate investment of time
and effort, replicate the results in the paper, as well as ask
additional questions.

Alas, reproducibility is not necessarily the norm in the relevant
literature.

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons {[}e.g., 52{]}. But
in many other cases the data are publicly available, which in fact has
commonly been the case with population-level COVID-19 information.
Typically the provenance of the data is documented, but in numerous
studies the data themselves are not shared {[}25,33,41,44,51,53--56{]}.
As any researcher can attest, whether a graduate student or a seasoned
scientist, collecting, organizing, and preparing data for a project can
take a substantial amount of time. Pointing to the sources of data, even
when these sources are public, is a small step towards
reproducibility-but only a very small one. Faced with the prospect of
having to recreate a data set from raw sources is probably sufficient to
dissuade all but the most dedicated (or stubborn) researcher from
independent verification. This is true even if part of the data are
shared {[}e.g., 46{]}. In other cases, data are shared, but the
processes followed in the preparation of the data are not fully
documented {[}48,57{]}. These processes matter, as shown by the errors
in the spreadsheets of Reinhart and Rogoff {[}16{]} and the data of
biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis {[}43,58,59{]}. Making code
available only ``on demand'' {[}e.g., 60{]} is an unnecessary barrier
when most journals offer the facility to share supplemental materials
online. Then there are those papers that more closely comply with
reproducibility standards, and share well-documented processes and data,
as well as the code used in any analyses reported {[}19,32,37,49,61{]}.

In the following sections, the analysis of RWN is replicated, some
relevant questions from the perspective of a spatial modeler are asked,
and the data are reanalyzed.

@@

SWN examined the association between the basic reproductive number of
COVID-19 and population density. The basic reproductive number \(R_0\)
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, \(R_0\)
measures how many new infections each infections begets. Infectious
disease outbreaks generally tend to die out when \(R_0<1\), and to grow
when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
number of cases to be able to assume that there is community
transmission of the pathogen. Accordingly, SNW based their analysis only
on counties that had at least 25 cases or more at the end of the
exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
Rico, and island territories.

\begin{figure}
\centering
\includegraphics{R0-Density-Reanalysis_files/figure-latex/R0-map-1.pdf}
\caption{\label{fig:R0-map}Basic reproductive rate in US counties
(Alaska, Hawaii, Puerto Rico, and territories not shown).}
\end{figure}

SWN used for their multivariate analysis mixed linear models. This is a
reasonable modelling choice: \(R_0\) is an interval-ratio variable that
is suitably modeled using linear regression; further, as SWN note there
is a likelihood that the process in not independent ``among counties
within each state, potentially due to variable resource allocation and
differing health systems across states'' (p.~3). A mixed linear model
accounts for this by introducing random components (in the case of SNW,
random intercepts at the state level). SWN estimated various models with
different combinations of variables, including median household income
and prevalence of travel by private transportation. These are sensible
controls, given potential variations in behavior: people in more
affluent counties may have greater opportunities to work from home, and
use of private transportation reduces contact with strangers. Moreover,
they also conducted various sensitivity analyses. After these efforts,
SNW conclude that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

Table \ref{tab:swn-results} replicates Table 1 of SWN (p.~6).

\begin{table}

\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Replicating SWN: Models 1-3}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Variable & beta & 95\% CI & beta & 95\% CI & beta & 95\% CI\\
\midrule
Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
\addlinespace
Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
\bottomrule
\end{tabular}}
\end{table}
